%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
%\usepackage{natbib}
\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\Richard}[1]{{\color{drab}Richard: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1}
\newcommand{\refb}[1]{\textbf{R2:} #1}
\newcommand{\editor}[1]{\textbf{Editor:} #1}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}}


\begin{document}
\section*{Response to reviewers}

\subsection*{Associate Editor: Martin, Darren}

\response{
Thank you very much for handling our manuscript and providing these two constructive reviews that have helped us to improve the paper. Please accept our apologies for the delayed resubmission. My two co-authors have moved one (more than a year ago) and I have been caught up in SARS-CoV-2 related work for much of the year.

We realized that our original manuscript was misleading to people familiar with complex mixture models that partition the sequence data into categories with similar properties. We now make it very explicit that this is not what we study -- instead, we study the extent to which we can estimate specific properties of every column of an alignment. Many of the comments by Reviewer 2 suggest that they were expecting an analysis of partition models while in fact, we studied the identifiability of parameters describing selection at every single site in the sequence. Such inferences can complement deep-mutational scanning techniques and become increasingly feasible as more and more diverse sequences of RNA viruses are available. We, therefore, think that the paper is of interest to the readership of Virus Evolution.
}

\editor{This seems like a good paper but in its current state virus evolution may not be a very good fit for it.  I think though that if you follow reviewer 2's recommendations (especially with respect to providing practical advice to virologists) it should be relatively straightforward to make the paper a bit more accessible to virus evolution's readership.}

\response{
We have rewritten large parts of the manuscript to make our work more accessible and relevant to the readership of Virus Evolution.
Since our primary objective is to learn about the biology of virus from sequence diversity, rather than develop methods for phylogenetic inference, we believe Virus Evolution and not Systematic Biology is an appropriate venue for this work.
The referee reports were very useful in rewriting and refocussing our work.

We provide a point-by-point response below.
}


\subsection*{Reviewer: 1}

\refa{Puller, Sagulenko, and Neher present a systematic look at the strengths and weaknesses of site-specific substitution models. They propose a new method for inferring site-specific parameters, investigate the limitations of the model, and applied it to within-in host and global HIV sequences. This work is clear, methodical, and will be of great interest to those who work on site-specific models. I recommend this work for publication. I have also listed some comments below but do not see any of them critical to acceptance.}

\response{Thank you very much for this assessment.}

\refa{- The authors mention that other methods to infer site-specific parameters do exist. How does their new algorithm compare in terms of speed or accuracy?}

\response{The most obvious software to compare our results to would be \texttt{swmutsel}.
However, \texttt{swmutsel} specifically models sequences are amino acides and their codons, something that isn't currenly implemented in treetime. In terms of run-time, swmutsel takes hours where the site specific inference by treetime takes minutes.
But we are afraid the results are not immediately comparable.
We have hence refrained from attempting a comparision.
}


\refa{- For the simulations, the author draw site-specific preferences from a Dirichlet distribution with $\alpha$ equal to 0.2 or 0.5. It would be nice to present a more interpretable value of the preferences as well, such as entropy or effective number.}

\response{
Thank you for this suggestion.
For amino acid alphabets, $\alpha=0.2$ and $\alpha=0.5$ correspond to an average entropy of $1.8$ or $2.3$ (in natural log units), respectively.
Calculating the effective number of states as $\left(\sum_i p_i^2\right)^{-1}$ yields 4.7 or 7.8 effective states.
We added th the latter to the manuscript.}


\refa{- Fig 5 and Fig S4 are both over-plotted and making the points transparent or adding histograms would help with interpretability. In Fig 1, the differences in linetype should be included in the legend. In Fig 2, it would be easier to keep the models straight if the lines were annotated on the graph directly instead of relying exclusively on color.}

\response{
We are grateful for these suggestions to improve the figures:
\begin{itemize}
    \item we reduced marker size and opacity in Fig 5 and Fig S4 to reduce overplotting. We also experimented with two dimensional histograms, but bins at the border of the domain (almost invariable sites) dominate to a degree that it is hard to see differences in the remainder of the histogram.
    \item we added second legend to figure 1 that maps line style to samples size. We also removed the curves corresponding to naive estimates by alignment frequencies. These are also include in Fig 2 and hence not needed here.
    \item labeling individual lines in Fig 2 proved a bit tricky, but we added markers to the lines such they can be distinguished without reference to color.
\end{itemize}
}


\subsection*{Reviewer: 2}
\response{
We thank the reviewer for their detailed review of our manuscript and the many constructive comments.
This review as revealed that we should have been much more explicit about the relation of our manuscript to partition models commonly used to model site-heterogeneity of rates or preferences.
Instead of partitioning the sequence data into a small number of categories, we investigate to what extent we can learn parameters that describe specific selective pressures at every column in the alignment.
The primary objective is to learn about the function of the sequence, not phylogenetic inference.
}

\refb{Abstract, line 19: I would phrase this more carefully, as I personally feel that site-specific substitution models are in fact quite popular (with indeed their own site-specific evolutionary rate being assigned/assumed as well) and easy to set up in many inference packages.}

\response{Thank you for this note. We realized that the intention and model class studied in our manuscript wasn't clear enough. We now emphasize that we study whether models with different state preferences at every site can be accurately estimated.
We now point out in multiple places that site heterogenous models with partitions are a common way to include biological complexity and that model selection has to be used to determine model complexity.
}

\refb{Abstract, line 21: It needs to be pointed out that a careful model selection strategy should already be part of a standard phylogenetic/phylodynamic data analysis, where indeed site-specific models are being included in the mix. That’s certainly how I do things, but then I don’t see what would be different with what the authors present here (more to be uncovered as I read through the manuscript obviously, but would be good to point out how the suggested approach will differ from such a model selection study).}

\response{
We now point out our objective more clearly: Our primary goal is to infer site specific preferences to learn about the selection pressures operating on the genome, not phylogenetic inference.
Specifically, we ask how much data is required to estimate state preferences at every site, how uncertainty in topology affects the estimates, and which aspects of the model are well or poorly constraint.
As such, the problem is distinct (but related) to a model selection problem.
}

\refb{Abstract, line 22: why ‘approximate’? How does this differ from ‘regular’ ML?}

\response{The scheme is approximate, in that the parameters that maximize the likelihood are found through iterative solution of extremal conditions obtained through an approximation of the full likelihood. }

\refb{Abstract, lines 24-25: I assume it will be pointed out that the BEAST/BEAST2 packages do allow for site-specific models to be used (with for example a simple button that allows to select the SRD06 model, of which the paper should be cited)?}

\response{We now discuss site-heterogenous partition models carefully in the main text.}


\refb{Lines 120-122: This brings to mind a recent paper by Gascuel and Steel ‘A Darwinian Uncertainty Principle’ in which the issue of jointly estimating the ancestral states and the transitions of those states is investigated. Given the parameters the authors are interested in here and their assessment through simulations, it seems to me that the authors need to clarify the novelty of their investigations and how theirs sets them apart from those investigated by Gascuel and Steel (actually Figure 1 reminds me of Figure 2 in the paper by Gascuel and Steel).}

\response{The paper by Gascuel and Steel (which we were unaware of) discusses indeed similar topics as we do. Gascuel and Steel investigate whether the state of the root of a character can be inferred jointly with the equilibrium probabilities or the transition matrix. Consistent with our results, they find that this is possible for Yule trees.
Here, we are mostly concerned with estimating the model, not the root state, and treat ancestral states as nuissance parameters that at least in principle can be integrated out.
In constrast to Gascuel and Steel, we consider sequence evolution rather than the evolution of a single character which allows our model to share the symmetric transmission matrix across sites.
It also allows us to investigate the extent to which branch lengths can be estimated along with the model.

We now discuss our results in the light of the results by Gascuel and Steel and thank the reviewer for pointing out this work to us.
G\&S's work is mentioned in the section discussing the accuracy of inferences and the discussion.
}

\refb{Line 141: True, and this deserves a lot more explanation in the manuscript, i.e. how far should one typically push the site-specific assumptions? The authors need to make it explicitly clear if they plan to estimate (and assess) one or more site-specific parameters for each site, or whether some of these parameters are shared across categories of sites as is most commonly done when dealing with data partitions (such as partitioning by codon position). If each site gets its own parameter(s), then overfitting seems to be a very real issue; but what about other cases such as pre-defined partitions? Or what about drawing certain parameters from a certain distribution or a hierarchical phylogenetic model, which could very much aid with overfitting issues by sharing data?}

\response{We indeed investigate models with multiple free parameters per site -- in contrast to category models.
In our model the symmetric substitution matrix is shared across sites, but each site has has its own rate and frequencies.
Much of the remainder of the manuscript is devoted to the issue of identifiablility and over-fitting.
As data sets grow in size, such site specific parameters become increasingly identifiable and this is focus of this study.
The properties of category models have for example been studied systematically by Kainer and Lanfear.
We now point out the risk of overfitting more clearly at the beginning of the section ``Accuracy of model inferences'' and emphasize that this is one of the research questions we want to address: How much data do you need to estimate such parameters?
}


\refb{Line 150: `data sets'}
\response{fixed, thanks.}

\refb{Line 157: 'tree'}
\response{fixed, thanks.}

\refb{Lines 204-208: Again, this seems related to Gascuel and Steel.}

\response{Indeed, Gascuel and Steel show a similar graph for the joint estimation of ancestral state of a single discrete character and the model of its evolution. We now point out this analogy.
The connection to Gascuel and Steel is particularly apt when discussion the inability to jointly infer the ancestral state and the model.
We also draw attention to the fact that accurate model estimation is possible for known trees, consistent with the results by G\&S that joint ancestral reconstruction and model inference is possible for Yule trees.
}

\refb{In my opinion, the ‘Accuracy of inferences’ section is hard to digest, mostly because of its technicality and -- in my personal opinion -- due to the fact that the figures (specifically figures 1 and 2) aren’t very intuitive so I would suggest the authors to rework both the use of the Y-axis (if the authors want to discuss accuracy in the text, then the graphs should be going up in the figures as well), as well as the legends within the figures, which also aren’t self-explanatory and require to go through the text in detail.}

\response{We have rewritten a good part of this section. We now point out that we first investigate the performance of the inference scheme with known tree and ancestral states and then investigate the feasibility of estimating tree, ancestral states, and model together.
Reviewer 1 also had suggestions to improve the figures and we have added more information to the legends and made it easier to differentiate between different line.
We adjusted to text to more explicitly point out that the graphs show squared error, i.e., a quantity that decreases with increasing accuracy.
}

\refb{Lines 239-240: I’m not particularly convinced that uniform frequencies are still the norm ...}

\response{We are not making a statement here on what is the norm and we are aware that complex models with multiple partitions are common. However, many large phylogenetic analysis of viruses use models that account for rate heterogeneity among sites, but use a common matrix of relative substitution rates for all site (PMID 25278604 for example). }

\refb{Line 308: I fear this example analysis is quite (or too) technical and am not sure this will appeal to typical readers of Virus Evolution. Overall, the manuscript would be more suited for a more methodological journal like Systematic Biology or Bioinformatics.}

\response{We have made an effort to make this section more accessible, but we think it is an important part of the manuscript and one of the reasons we think Virus Evolution is an appropriate journal.
This section explores the ability to estimate site specific preferences from large and diverse alignments of RNA viruses.
Diverse viruses like HIV are one place were data to apply such methods are available and the interpretation of conservation and viral fitness is of interest to many readers of Virus Evolution.
Furthermore, our work has implications for dating of divergence times of RNA viruses.
We don't have solutions for the latter, but feel that it is still important to point them out.
}


\refb{But what I’m missing the most from the results and discussion sections of the manuscript, is clear advice on how to perform analyses on real data, what to look out for and how to potentially remedy the problem. The manuscript is too technical in that regard, and while assessing accuracy problems is important, a step-by-step approach for empirical data sets- to potentially/hopefully mitigate the problems mention in this manuscript - is even more important. It would be most helpful if the authors comment on the typical workflow, i.e. running ModelTest or something similar (PartitionFinder, automatic/included model selection in IQ-TREE, ...) and then sticking with the model found to infer the phylogeny. How often would one get in trouble with such an approach? And if problems arise, how severe are they? Could the clustering potentially be impacted or does the problem stay limited to certain branches (of certain length)? Or is the tree as a whole deemed older/younger than it really is, i.e. does the underestimation propagate through the tree all the way to the root or not?}

\response{
We hope that the revision makes now clear that phylogenetic inference is not the focus of the manuscript, but that we instead study how properties of sites can be learned from sequence diversity.
Such inference from large viral data sets is increasingly feasible as data set size increases and complements deep mutations scanning approaches.
While we identify problems for phylogenetic inference associated with strong variation model variation along the sequence, we don't have a solution.
The site specific models we estimate could serve as input for ParitionFinder or the "-MF+MERGE" feature of IQ-tree.
}


\refb{In a way, I’m not too sure what the conclusions from Figure 3 are in practice? Everything except for the true model results in underestimation, even the site-specific model in TreeTime. From a practical point of view, what could be done to fix the underestimation problem? Would some form of post-hoc correction be advised, but how would one know which branches to correct and which to leave along? In that regard, could thorough knowledge of the data provide any guidance; but even then, and related to the HIV example (and other virus examples), inaccurate branch lengths of 100 years or more are most likely not a major concern.}

\response{In essence, Fig 3 shows that branch length underestimation caused by extensive variation of rates and preferences is very hard to avoid.
As outlined above, we don't really have a good fix for that, but we investigate the causes in Fig 4.
That said, the main result of the paper is that site-specific frequencies can be accurately estimated.
The inability is estimating the length of long branches in RNA viruses is a serious concern for dating zoonosis and exploring the history of viruses causing measles, dengue, lassa etc.
This problem is clearly apparent when trying to date the common ancestor of SIVs.
Molecular clock approaches suggest an ancestor a few hundred years in the past, while viruses in island populations show that the common ancestor is much more ancient.
}



\refb{The manuscript deals with estimates (i.e. trees) that were obtained using IQ-TREE and FastTree (which produce unrooted trees and don’t employ the concept of a molecular clock), which were subsequently fed into TreeTime which employs the provided trees as is. This makes me wonder if the conclusions from the manuscript hold up in frameworks that i) employ a molecular clock directly in their estimations, and ii) do not employ a fixed topology but rather explore a distribution of trees as typically done in Bayesian phylogenetic inference. I’m not suggesting the authors do the same experiments in a Bayesian framework but their take on this could be interesting for the discussion section.}

\response{
We are afraid there is a misunderstanding here.
Our manuscript infers GTR models and the root does not matter.
We use root-to-tip distances in mid-point-rooted trees as a measure of divergence, but the root is no special place otherwise.
The only place where time comes in are the HIV examples, but we only use time as a proxy for divergence via the known evolutionary rate and don't do any molecular clock analysis.
All we want to do is to point out that branch length under-estimation potentially has consequences for dating rapidly evolving RNA viruses.}


\refb{Lines 651-653: I’m going to assume that most readers, like myself, will not be very familiar with the work of Halpern and Bruno (1998). Hence these sentences will need to provide more information regarding the following statement: ‘only allow for site-specific rates and equilibrium frequencies while using the same transition matrix for every site.’ Regarding the site-specific (evolutionary) rates, there may be confusion as to whether the process meant is actually among-site rate variation (typically modelled as a discretized gamma distribution according to various paper of Ziheng Yang), or relative rates between positions which can be superimposed on the former model. I’m going to assume the latter, but in that case it’s not clear whether for each site a unique rate will be estimated (if so, that’s a lot of additional parameters and this needs to be discussed) or whether classes of sites will be imposed (the three different codon positions for example) and that for only each of these classes a unique rate will be estimated? Essentially, I have the same question regarding the equilibrium frequencies, as it’s not clear from the methods section (but I can guess from the simulated data section; still, the authors should be more explicit here).}

\response{
The points raised by this reviewer have made it clear to us that the description of our basic motivation and model set-up was not clear enough. In response to this, we have rewritten several parts of the manuscript and point out explicitly in multiple places that we do not consider partition models but investigate to what extend models in which every site has their own set of of frequencies and an overall rate can be estimated. }


\refb{Implementation: this type of modelling is not new, although it surely is a useful -- and needed -- extension to TreeTime. It’s unclear at this point whether the transition matrix will also be (potentially) site-specific or not, and I encourage the authors to mention this explicitly in the implementation paragraph.}

\response{We currently implement the model as specified in Equation 1. We now mention this explicitly and point out that site specific $W_{ij}$ are not currently implemented.}


\refb{Line 681: This makes it sounds like either two data partitions will be simulated that will either be concatenated or the contents will be scrambled to create one alignment of length L=2000 that is then a mixture of the different (site-specific) processes. But I have the feeling that this is not what the authors are after; can this be more explicitly explained?}

\response{The reviewer is correct, we don't have two data partitions of length 1000.
Every site has its own set of frequencies and an overall rate.
Instead, we produce hierarchical data sets that allow us to explore the effect of different aspects of the problem (tree, model realization, realization of the sequence evolution given the model) on the inferences.
We added sentence to make this clear. }


\refb{Lines 690-691: this may be confusing for readers, i.e. why is TreeTime not mentioned here? It should be pointed out then that a tree topology is required to be provided to TreeTime.}

\response{We added `using TreeTime' to the sentence.}

\refb{Line 708: ‘site-specific GTR inference’, will the site-specific processes be bound using a (Gamma) distribution as was explained/used in the simulation section?}

\response{The site specific models were estimated using iterative approach described in Eq. 2 which effectively regularizes the problem with a Dirichlet prior. }


\end{document}